Investigating soundscapes perception through acoustic scenes simulation

### 2018-07-16 v2/responseToReviewers.tex

\section{Answer to Reviewer 1}

\begin{enumerate}

\item \emph{The utility of this tool beyond the experiments presented here is not apparent because the flexibility of the system was not demonstrated, for example:
- SimScene itself was only used in the first experiment, and other applications of it were not considered. Design choices vary widely between researchers and domains, meaning that the specific instantiations here may not be generally useful.}

$\rightarrow$ We definitely agree. As now discussed in more detail in Section 6 of the manuscript, the proposed software tool is not designed for genericity but we believe that the proposed specific design strategies and use of recent software toolkits can be considered as inspirations for desiging other experimental protocols questionning soundscape perception.

### 2018-07-16 lafaySimulatedScenePerception.tex
integrates jfp comments

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used by~\cite{Salamon14}. Only three levels of abstraction are considered for the texture sounds.

@@

As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more toroughly described by~\cite{rossignol2015simscene}.

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to simulate urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}). Each of them has to create two sound environments: one ideal~/ pleasant, and the other non ideal~/ unpleasant.
\item \emph{experiment 2.a (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal or non ideal. Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non ideal) thanks to a finer quantification of the pleasantness of the scene;
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or non ideal) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non ideal scene.
\end{enumerate}
\end{itemize}

@@

\begin{itemize}
\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item second simulation: create a \textbf{plausible urban soundscape} which is non ideal, according to you (where you would not like to live);
\end{itemize}

@@

\begin{enumerate}
\item Simulation, where the user is asked to:
  \begin{itemize}
  \item  select sound classes,
  \item  give each of them a name,
  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
  \end{itemize}
\item Feedback : writing of a free form comment about the composed soundscape.
\end{enumerate}

@@

All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and are not allowed to talk to each other during the experiment.

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non ideal scenes). the 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non ideal scenes are first sequenced to allow the subjects to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account. Each participant evaluates all the sound scenes.

@@

All the subjects perform the experiment simultaneously. They are not allowed to talk to each other during the experiment.

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for ideal scenes and non ideal scenes.

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non ideal scenes during the experiment 1.a. To do so, we consider if there exist significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non ideal scenes. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non ideal scenes. The significance is measured with a two-sample Student test.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k=\lbrace ideal, non ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between ideal and non ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

First, in order to ensure the coherence of the data, we check that none of the non ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non ideal scenes.

@@

\begin{figure}[t]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure}

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between ideal and non ideal scenes ($L$: $p<0.01$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the non ideal scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non ideal scenes. This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, level alone is not sufficient to fully differentiate between ideal and non ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & ideal scenes          & non ideal scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.

@@

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non ideal scenes.

@@

\begin{itemize}
\item allows us to differentiate between ideal and non ideal scenes,
\item characterizes precisely the perceived pleasantness for non ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (ideal scene).
\end{itemize}

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non ideal scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non ideal scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non ideal scenes, texture levels do not bring much information, whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non ideal), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.

@@

  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
                             \caption{Proportion of simulated scenes (i: ideal, ni: non ideal) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$. }\label{fig:soundsource}
\end{figure}

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (ideal or non ideal). For the 36 ideal and 36 non ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a noticeable difference in terms of class choices between the ideal and non ideal scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non ideal scenes. These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.

@@

Nevertheless, we notice some differences with results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.

To a certain extent, our results contradict this result. In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non ideal scenes. Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the ideal scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. However, \emph{public transport} classes are also strongly present in the non ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between ideal or non ideal scenes. We now investigate whether some of the sound classes are specific to a given environment. For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

To conclude, the level of ni- markers has a negative influence on the pleasantness for the non ideal scenes. On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.

>>>>

\begin{enumerate}
\item \emph{experiment 1.a, simulation}:  the subjects use the soundscape simulator to produce ideal / non-ideal soundscapes that are considered as material for the following experiments; The resulting waveforms are available for download \url{https://archive.org/details/soundSimulatedUrbanScene}.
\item \emph{experiment 1.b, pleasantness evaluation}: the subjects judge the pleasantness of the simulated scenes on a semantic scale;
\item \emph{experiment 2, pleasantness evaluation after modification of the scenes}: the subjects judge the pleasantness of the simulated scenes on a semantic scale as in 1.b, but some scenes are modified beforehand, \ie~some specific sounds classes that are identified as having a significant impact on perceived pleasantness are removed.
\end{enumerate}

@@

Accounting for the previously detailed perceptual matters, two taxonomies are built, one for event sounds and one for texture sounds. Four levels of abstraction are considered from the most generic (level 0) to the most specific classes (level 2), leading to a taxonomy close to the one used in~\cite{Salamon14}, see Figure~\ref{fig:taxonomieEb}. Only three levels of abstraction are considered for the texture sounds, see Figure~\ref{fig:taxonomieT}.

@@

As such, the underlying model of the scene is a sum of sound sources. The simulation interface is more thoroughly described by~\cite{rossignol2015simscene}.

@@

\begin{itemize}
\item \emph{experiment 1.a (simulation)}: during this experiment, subjects are asked to create simulated urban sound environments using \emph{Simscene} (see Section~\ref{sec:simulator}). Each of them has to create two sound environments: one ideal~/ pleasant, and the other non-ideal~/ unpleasant.
\item \emph{experiment 1.b (evaluation)}: after the simulation phase, only a binary information on the pleasantness property of the respective scenes is available: respectively ideal or non-ideal. Furthermore, this information is given by the creator of the scene. The second experimental step aims at investigating more deeply and more broadly our knowledge on the pleasantness of the simulated scenes. For that, a second group of subjects is asked to evaluate the pleasantness of each scene produced during (1.a), on a semantic scale. This experiment has two goals:
\begin{enumerate}
\item to evaluate more precisely the respective influence of the various sources composing the scenes on the pleasantness  (ideal or non-ideal) thanks to a finer quantification of the pleasantness of the scene;
\item to detect the presence of \emph{outliers} or ambiguous scenes. Indeed, throughout our analyses, the predefined hedonic properties of the scenes (ideal or non-ideal) are used as reference. We thus need to ensure beforehand that no ambiguity exists between extreme cases of ideal and non-ideal scenes, \ie~that the least pleasant ideal scene remains more highly rated on average than the most pleasant non-ideal scene.
\end{enumerate}
\end{itemize}

@@

\begin{itemize}
\item  first simulation: create a \textbf{plausible urban soundscape} which is ideal, according to you (where you would like to live);
\item second simulation: create a \textbf{plausible urban soundscape} which is non-ideal, according to you (where you would not like to live);
\end{itemize}

@@

\begin{enumerate}
\item Simulation, where the user is asked to:
  \begin{itemize}
  \item  select sound classes,
  \item  give to each of them a name,
  \item  set the parameters of the tracks related to the selected sound classes of sounds, see Section~\ref{sec:simscene_parametre}.
  \end{itemize}
\item Feedback : writing of a free form comment about the composed soundscape.
\end{enumerate}

@@

All the subjects performed the experiment at the same time. They were equally distributed in three identical quiet rooms, and were not allowed to talk to each other during the experiment.

@@

Among the 44 subjects, 40 succeeded, producing at the end 80 simulated sound scenes (40 ideal scenes, 40 non-ideal scenes). The 4 other subjects were excluded from the process due to a lack of understanding of the instructions or failure to respect them, or for exceeding the maximum duration allowed to perform the experiment.  The software platform used for the experiment, the parametrization of the software platform for each generated scene, as well as a 2 dimensional projection of the resulting scenes are available \url{http://soundthings.org/research/urbanSoundScape/XP2014}.

@@

The assessment is done with a 7-point bipolar semantic scale going from -3 (non-ideal / unpleasant) to +3 (ideal / pleasant). Before evaluating a scene, subjects must listen to the first 20 seconds of the stimuli. After the evaluation, they are free to continue to the next scene.

@@

For each participant, sound scenes are played in a quasi random order. 5 ideal scenes and 5 non-ideal scenes are first sequenced to allow the subjects to calibrate their scores. These first 10 scenes are played back again at the end of the experiment. Only the last evaluations are taken into account. Each participant evaluates all the sound scenes.

@@

All the subjects perform the experiment simultaneously. They were not allowed to talk to each other during the experiment.

@@

\begin{itemize}
\item \emph{perceptual features}: the perceived pleasantness of the composed scene, assessed on a 7-point bipolar semantic scale. $\mathcal{A}_{scene}$ is the average pleasantness for one scene, as the average of the marks of all the subjects. $\mathcal{A}_{subject}$ is the average pleasantness for each participant, computed as the average of all the score given by each participant. Given the low number of subjects in experiment 1.b, we choose not to normalize the pleasantness scores. $\mathcal{A}_{subject}$ is computed for ideal scenes and non-ideal scenes.

@@

\item \emph{Analysis of perceived pleasantness}: the goal is to evaluate whether the perceived pleasantness is in accordance with the pleasantness label given by the creators of the  ideal and non-ideal scenes during the experiment 1.a. To do so, we consider if there exists significant differences between the average $\mathcal{A}_{scene}$ and the average $\mathcal{A}_{subject}$ of the ideal and non-ideal scenes. The significance is evaluated by a two-sample Student test for $\mathcal{A}_{scene}$ and by a paired-sample Student test for $\mathcal{A}_{subject}$.

@@

\item \emph{Analysis of sound levels}: the goal is to evaluate whether the sound levels differ between the ideal and non-ideal scenes. The significance is measured with a two-sample Student test.

@@

\item \emph{Analysis of semantic features}: the goal is to evaluate if specific classes are more frequently used to simulate a given type of environment (ideal or non-ideal). To do so, a V-test is considered, see~\citep{rakotomalala2008tvpercent}. With $c$ being the total number of classes used for both types of environment, $c_k$ the number of classes used for a given type of environment $k$ ($k \in \lbrace ideal, non-ideal\rbrace$), $c_j$ the number of times a class $j$ has been used, and $c_{jk}$ the number of time a class $j$ has been used for a given type of environment $k$, the V-test evaluates the null hypothesis that the ratio $\frac{c_{jk}}{c}$ is not significantly different from the ratio $\frac{c_{jk}}{c_k}$. For each class $j$, and each environment type $k$, an approximation of the statistical value $V_{jk}$ is computed as follows:

@@

where $\bigoplus$ is the \emph{exclusive-or} operator. Two scenes having similar source compositions will be close in such space. Using the Hamming distance allows us to take into account equally the presence and absence of classes. In order to measure the intrinsic ability of the space to discriminate between ideal and non-ideal scenes, we use a ranking metric named the precision at rank $k$ ($P@k$). The $P@k$ computes the precision obtained after the $k$ closest items with respect to a given seed item have been found. Formally, for each $s_i$ scene (considered as seed), we compute the proportion of $s_j$ scenes in the $k$ nearest neighbors of $s_i$ that share the same label as $s_i$. The $P@k$ is then the average of this ratio for all the items considered as search seeds.

@@

First, in order to ensure the coherence of the data, we check that none of the non-ideal scenes gets a $\mathcal{A}_{scene}$ higher than one of a ideal scene. 4 non-ideal scenes do not fulfill that constraint: they are thus removed, together with their corresponding ideal scenes. As a consequence, 36 ideal scenes and 36 non-ideal scenes remain for analysis. Second, we verify that subjects really perceived a difference in terms of pleasantness between ideal and non-ideal scenes. For that, we investigate the mean pleasantness score for each participant $\mathcal{A}_{subject}$, computed separately for each type of environment. It indeed appears that the ideal scenes were perceived significantly more pleasant ($p < 0.01$) than the non-ideal scenes for all the participants.

@@

\begin{figure}[t]
		\myfloatalign
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_1}\label{fig:soundlevela}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_3}\label{fig:soundlevelb}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_5}\label{fig:soundlevelc}}\par
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_2}\label{fig:soundleveld}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_4}\label{fig:soundlevele}}
        \subfloat[]
        {\includegraphics[width=.33\linewidth]{gfx/xp_soundlevel_6}\label{fig:soundlevelf}}
       \caption{Distributions of the sound levels $L$ (a, d), $L(E)$ (b, e) et $L(T)$ (c, f), with respect to scene type (i: ideal, ni: non-ideal) (a, b, c) and perceived pleasantness $\mathcal{A}_{scene}$ of experiments 1.b (d, e, f).}
\end{figure}

@@

First, our analysis focuses on the sound levels. Figures ~\ref{fig:soundlevela},~\ref{fig:soundlevelb} and~\ref{fig:soundlevelc} respectively depict the distributions of levels $L$, $L(E)$ and $L(T)$. There is a significant difference in terms of sound levels between ideal and non-ideal scenes ($L$: $p<0.01$). This difference is significant for events ($L(E)$: $p<0.01$, mean deviation: -7 $dB$) and for textures ($L(T)$: $p<0.01$, mean deviation: -6 $dB$).

@@

As expected, the sound level of the sources is indeed a pleasantness indicator, as the non-ideal scenes tend to be louder. This results is also an outcome of a large number of related works. We also notice that this difference of sound levels is significant for both events and textures.

@@

It appears that the biggest influence on the global sound levels comes from the events, the difference between $L$ and $L(E)$ being only 1 $dB$ for ideal and non-ideal scenes. This observation is in agreement with the results obtained by~\cite{kuwano_memory_2003}. During their experiment, the authors ask their subjects to assess a set of soundscapes at a global level and then to do the same judgment at the time when they detect a sound source. The study shows that there is no significant difference between global and averaged instantaneous judgments. In our case, the result can be interpreted as if the subjects had unconsciously integrated this perceptual reality when composing the scenes, by allocating most of the global sound levels to well identified and relatively short sounds, \ie~the events.

@@

However, level alone is not sufficient to fully differentiate between ideal and non-ideal scenes. In fact, $20\%$ of the ideal scenes have a sound levels higher than the lowest level of the non-ideal scenes, while there is no overlap when considering the perceived pleasantness $\mathcal{A}_{scene}$.

@@

\begin{table}[t]
\centering
\begin{tabular}{l c c c}
               & all scenes                & ideal scenes          & non-ideal scenes    \\
\hline
$L$            & \textbf{-0.77} ($p<0.01$) & -0.32 ($p=0.06$)  & \textbf{-0.78} ($p<0.01$)\\
$L(E)$         & \textbf{-0.75} ($p<0.01$) & -0.20 ($p=0.24$)  & \textbf{-0.75} ($p<0.01$)\\
$L(T)$         & \textbf{-0.53} ($p<0.01$) & -0.33 ($p=0.05$)  &  -0.00 ($p=0.99$) \\
\hline
\end{tabular}
\vspace{0.5mm}
\caption{Linear correlation coefficients computed between mean perceived pleasantness $\mathcal{A}_{scene}$ of experiment 1.b and sound levels.}
\label{tab:corrStructA}
\end{table}

@@

In this section, more detailed relationships that could exist between sound levels and perceived pleasantness are investigated. Contrary to the previous test, we do not limit ourselves to a binary ideal scenes \emph{vs.} non-ideal scenes disctinction: we consider here the mean pleasantness $\mathcal{A}_{scene}$ as the perceptual feature. The aim is to investigate the level of correlation between sound levels and $\mathcal{A}_{scene}$. The linear correlation coefficients computed between $\mathcal{A}_{scene}$ \vs~$L$, $L(E)$, $L(T)$ are shown in Table~\ref{tab:corrStructA}. Relationships between $\mathcal{A}_{scene}$ and the sound levels are depicted in Figure~\ref{fig:soundleveld},~\ref{fig:soundlevele} and~\ref{fig:soundlevelf}.

@@

Concerning $L$, a strong negative correlation with $\mathcal{A}_{scene}$ is measured ($r=-0.77$, $p<0.01$), indicating that the higher the sound level is, the more unpleasant the scene is perceived. Nevertheless, Figure~\ref{fig:soundleveld} suggests that this relationship does not occur in the same way for ideal and non-ideal scenes. In fact, the correlation between $L$ and $\mathcal{A}_{scene}$, remains high for non-ideal scenes ($r=-0.78$, $p<0.01$), but is not significant ($p=0.06$) for ideal scenes.

@@

When considering the whole set of scenes, the fact that the level is indeed a good indicator of pleasantness can be explained by the fact that the ideal scenes tend to be softer than the non-ideal scenes, thus allowing us to extend artificially to the ideal scenes the negative correlation observed for the non-ideal scenes.

@@

\begin{itemize}
\item allows us to differentiate between ideal and non-ideal scenes,
\item characterizes precisely the perceived pleasantness for non-ideal scenes (unsurprisingly, an unpleasant scene gets all the more unpleasant as it gets louder),
\item is not a relevant feature for modeling the perceived pleasantness of an \textit{a priori} pleasant soundscape (ideal scene).
\end{itemize}

@@

The same conclusions can be drawn for $L(E)$, see~Figure~\ref{fig:soundlevele}. For $L(T)$, as shown on~Figure~\ref{fig:soundlevelf}, the moderate correlation observed for the whole set of scenes disappears when separate scenes are considered (ideal scenes: $r=-0.33$, $p=0.05$, non-ideal scenes: $r=-0.00$, $p=0.99$). Again, we believe that the negative correlation coming from the whole is an artifact due to the level difference between the two types of scenes (ideal scenes tend to be softer than non-ideal scenes). Thus, while sound event levels maintain a relative ability to predict the pleasantness of the non-ideal scenes, texture levels do not bring much information, whatever the type of environment is.

@@

Those first outcomes tend to show that 1) two modes of perception exist depending on the nature of the environment (ideal or non-ideal), and 2) each involves distinct, independent features. The fact that $L$ is not sufficient to characterize the pleasantness of the ideal scenes can lead us to conclude that all sound sources do not equally contribute to the perception of pleasantness. We thus put forward the hypothesis that only the level of some of them can influence this perception. In order to investigate further in that direction, we analyze in the next section the scenes from a semantic point of view, \ie~we take an interest in the nature of the sources they are made of.

@@

  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_2}\label{fig:soundsourceb}}
  \subfloat[]{\includegraphics[width=.5\linewidth]{gfx/xp1_class_3}\label{fig:soundsourcec}}
                             \caption{Proportion of simulated scenes (i: ideal, ni: non-ideal) involving a specific class of sounds: (a) event classes at an abstraction level $0+$, (b) texture classes at an abstraction level $0$, (c) event sub-classes at an abstraction level $1$ belonging to \emph{traffic} and \emph{public transportation} classes of the abstraction level $0$. }\label{fig:soundsource}
\end{figure}

@@

We study the composition of the scenes by counting the number of subjects who used a given class of sounds to simulate a given type of environment (ideal or non-ideal). For the 36 ideal and 36 non-ideal scenes considered, results are shown on Figure~\ref{fig:soundsourcea} for events and on Figure~\ref{fig:soundsourceb} for textures. For the sake of clarity, a transitional level of abstraction between level 0 and 1, named $0+$, is used to depict classes, see ~Figures~\ref{fig:taxonomieEa}, \ref{fig:taxonomieEb} and \ref{fig:taxonomieT}.

@@

We observe a noticeable difference in terms of class choices between the ideal and non-ideal scenes. The distribution of the classes is very similar to the one obtained in a related work on ideal urban soundscapes by~\cite{guastavino2006ideal}, \ie~on one hand, classes involving human presence and nature are prevailing in the ideal scenes, and on the other hand, classes involving mechanical sounds and/or public works are prevailing in the non-ideal scenes. These results confirm a fact previously observed by~\cite{raimbault2005urban,dubois2006cognitive}: the semantic nature of the sound sources play an important role in the assessment of the environment.

@@

Nevertheless, we notice some differences with the results of~\cite{guastavino2006ideal} which show that sounds of \emph{public transportation} are specific of ideal urban soundscapes. The authors interpret this by the fact that the perception of pleasantness is, among other things, due to socio-cultural factors. Thus, in our representation of the world, sounds of public transportation are positively connoted and tend to be more accepted than sounds of personal vehicles.

To a certain extent, our results contradict this result. In fact,  Figure~\ref{fig:soundsourcea} shows that \emph{public transportation} classes (\emph{bus} and \emph{train}, \cf~Figure~\ref{fig:soundsourcec}) have been used by the subjects for $28\%$ of the ideal scenes, and for $42\%$ of the non-ideal scenes. Those results do not question the fact that sounds of \emph{public transportation} are well accepted: $25\%$ of the subjects used \emph{bus} for the ideal scenes, a level that is comparable to the \emph{bike} class, and much higher than all the personal vehicles classes. However, \emph{public transport} classes are also strongly present in the non-ideal scenes, for instance more than \emph{light vehicle} or \emph{truck} classes. On the basis of our results, the public transportation class cannot be considered as typical of an ideal urban soundscape.

@@

We have shown that, from a qualitative point of view, the composition of the scenes in terms of sound sources differs between ideal or non-ideal scenes. We now investigate whether some of the sound classes are specific to a given environment. For that purpose, the V-test detailed in  Section~\ref{sec:xp1_dataAna} is considered separately for each abstraction level. Results are presented in Table~\ref{tab:markers}.

@@

To conclude, the level of ni- markers has a negative influence on the pleasantness for the non-ideal scenes. On the other hand, the level of the ideal markers does not influence the perceived pleasantness for the ideal scenes.
